Cross-validation of regression models: mixed model examples (edited v3)

### 2024-07-24 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
edit JSS paper

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical model objects. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly = TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times. There are also data management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provide greater detail on some topics that we do describe, such as extensions to the package.

@@

```{r Auto}
data("Auto", package = "ISLR2")
summary(Auto)
```
With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto", package="ISLR2")}.

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a). The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).

@@

cv(m.auto, confint = TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint=TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times=100} repetitions.

@@

cv.auto.loo <- cv(mlist, data = Auto, k = "loo")
cv.auto.loo[2]
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} object returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

@@

Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models. The HSB data are used by @FoxWeisberg:2019 [Sec. 7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.

@@

The four data sets differ in the magnitude of between-cluster variance compared to within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

@@

[^code-in-complements]: The code for this section is in the complementary materials for the paper. We used the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variance. We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs.

@@

```{r plot1, echo=FALSE, cache=TRUE}
#| out.width = "75%",
#| fig.height = 2.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the ratio of between-patient SD to within-patient SD. The ellipses are estimated 50% Gaussian concentration ellipses for each patient. The population between-patient regression line, $\\E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       ylab = list('symptoms (y)', cex = 0.7),
       xlab = list('dosage (x)', cex = 0.7),
       par.strip.text = list(cex = 0.7),
       scales = list(cex = 0.7, x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```

Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}. The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}. We thus fit four models to four data sets for a total of 16 models.

@@

```{r plot-fits-combined,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
levs3 <- levels(Data$panel2)
levs3 <- sub(':fixed','', levs3)
levs3 <- sub(':BLUPs',' ', levs3)
Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)

@@

Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between \code{y} and \code{x} in the model with an intercept and \code{x} as fixed-effect predictors, \code{~ 1 + x}. The inclusion of the contextual variable \code{xm} in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the \code{~ 1 + x} model in the two data sets with the smallest between-patient variances. For data with a small between-patient variance, fixed-effects predictions for the \code{~ 1 + x} model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variance is relatively small, but that cluster-based cross-validation will prefer the latter to the former. We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' data set.

The third model, \code{~ 1 + x + xm}, includes a contextual effect of \code{x}---that is, the cluster mean \code{xm}---along with \code{x} and the intercept in the fixed-effect part of the model, and with a random intercept. This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | patient)}, which is the model that generated the data.  The fit of the mixed model \code{~ 1 + x + xm} is consequently similar to that of a fixed-effects-only model with \code{x} and a categorical predictor for individual patients (i.e., \code{y ~ factor(patient) + x}, treating patients as a factor, and not shown here).

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.

@@

```{r cross-validation-data,include=FALSE,cache=TRUE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "75%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variance is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

As @HastieTibshiraniFriedman:2009 [Sec. 7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to specify or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.  Statistical modeling is partly a craft, and one could imagine applying that craft to successive partial data sets, each with a fold removed. The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets. Alternatively, if we're able to automate the process of model specification, then we can more realistically apply CV mechanically.

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression.:

```{r Auto-explore-scatterplot-matrix, echo=FALSE}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data."
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("AMER", "EUR", "JAP"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

@@

```{r Auto-explore-cr-plots, echo=FALSE}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Component + residual plots for the working model fit to the \\code{Auto} data."
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
```

```{r Auto-explore-show, eval=FALSE}
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("AMER", "EUR", "JAP"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

scatterplotMatrix(~ mpg + displacement + horsepower + weight 
                  + acceleration, 
                  smooth = list(spread = FALSE), data = Auto, pch = ".")
```

A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders}, ranging from 3 to 8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most these variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model, eval=FALSE}
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
```
The component + residual (C+R) plots in Figure \ref{fig:Auto-explore-cr-plots}, produced by the \code{cvPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model. The broken blue lines in the C+R plots show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and 0s: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda = powers[pred])
}
m <- update(m.auto, data = A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore-scatterplot-matrix}, on page \pageref{fig:Auto-explore-scatterplot-matrix}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix, echo=FALSE}
#| out.width = "95%",
#| fig.align = 'center',
#| fig.height = 5.5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data."
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread = FALSE), data = A, pch = ".")
```

```{r Auto-transformed-cr-plots, echo=FALSE}
#| out.width = "95%",
#| fig.align = 'center',
#| fig.height = 5.5,
#| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
```

```{r Auto-transformed-scatterplot-matrix-show, eval=FALSE}
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth = list(spread = FALSE), data = A, pch = ".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:
```{r Auto-CR-plots-transformed, eval=FALSE}
crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
```

@@

* \code{data}, the data set to which the model is fit.
* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC=FALSE}, which specifies use of the BIC for model selection.

>>>>

The \pkg{cv} package for \proglang{R} automates the process of cross-validation for standard \proglang{R} statistical models. The principal function in the packages, also named \code{cv()}, has methods for objects produced by a number of commonly employed regression-modeling functions, including those for mixed-effects models:
```{r cv-methods}
library("cv", quietly = TRUE)
methods("cv")
```
 The \code{cv()} function is introduced in the context of a preliminary example in Section \ref{preliminary-example-polynomial-regression} of the paper.

@@

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times. There are also data-management facilities in the package, such as coercing the objects produced by the \code{cv()} function into data frames for further analysis. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provide greater detail on some topics that we do describe, such as extensions to the package.

@@

```{r Auto}
data("Auto", package = "ISLR2")
summary(Auto)
```
With the exception of \code{origin} (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see \code{help("Auto",} \code{package = "ISLR2")}.

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a). The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the degree-5 polynomial may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).

@@

cv(m.auto, confint = TRUE)
```
The function reports the CV estimate of MSE, a bias-adjusted estimate of the MSE (the bias adjustment is explained in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}), and the MSE is also computed for the original, full-sample regression. Because the number of cases $n = 392 < 400$ for the \code{Auto} data, we set the argument \code{confint = TRUE} to obtain a confidence interval for the MSE, which proves to be quite wide.

@@

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the patient reader to redo this computation with the default \code{times = 100} repetitions.

@@

cv.auto.loo <- cv(mlist, data = Auto, k = "loo")
cv.auto.loo[2]
```
The \code{models()} function takes an arbitrary number of regression models as its arguments, which are optionally named, to create a \code{"modList"} object. Alternatively, as here, we can supply a pre-constructed, optionally named, list of models as the single argument to \code{models()}. To visualize the results, comparing the models, we invoke the \code{plot()} method for the \code{"cvModList"} objects returned by \code{cv()} (see Figure \ref{fig:polynomial-regression-CV-graph-2}):

@@

Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* ("HSB") survey have become a staple of the literature on mixed-effects models. The HSB data are used by @FoxWeisberg:2019 [Sec. 7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.

@@

The four data sets differ in the magnitude of between-cluster variation compared to within-cluster variation. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variation, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

@@

[^code-in-complements]: The code for this section is in the complementary materials for the paper. We used the \code{glmmTMB()} function in the \pkg{glmmTMB} package [@BrooksEtAl] for the models fit here because, in our experience, it is more likely to converge than functions in the \pkg{nlme} and the \pkg{lme4} packages for models with low between-cluster variation. We used the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for the graphs.

@@

```{r plot1, echo=FALSE, cache=TRUE}
#| out.width = "75%",
#| fig.height = 2.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variation. The labels above each panel show the ratio of between-patient SD to within-patient SD. The ellipses are estimated 50% Gaussian concentration ellipses for each patient. The population between-patient regression line, $\\E(y) = 10 + x$, is shown in each panel, where $y$ is symptoms and $x$ is dosage."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       ylab = list('symptoms (y)', cex = 0.7),
       xlab = list('dosage (x)', cex = 0.7),
       par.strip.text = list(cex = 0.7),
       scales = list(cex = 0.7, x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```

Using data like these, a researcher may attempt to obtain an estimate of the within-patient effect of dosage by fitting mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variation at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested in their fixed effects: (1) \code{~ 1}, intercept only; (2) \code{~ 1 + x}, intercept and effect of \code{x}; and (3) \code{~ 1 + x + xm}, intercept, effect of \code{x}, and a contextual variable, \code{xm}, consisting of the within-patient mean of \code{x}. The final model, (4) \code{~ 1 + I(x - xm)}, uses an intercept and the centered-within-group variable \code{x - xm}. We thus fit four models to four data sets for a total of 16 models.

@@

```{r plot-fits-combined,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient standard deviation ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses, along with the patient centroids."
levs3 <- levels(Data$panel2)
levs3 <- sub(':fixed','', levs3)
levs3 <- sub(':BLUPs',' ', levs3)
Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)

@@

Data sets with relatively low between-patient variation result in strong shrinkage of fixed-effects predictions, and also of BLUPs, towards the between-patient relationship between \code{y} and \code{x} in the model with an intercept and \code{x} as fixed-effect predictors, \code{~ 1 + x}. The inclusion of the contextual variable \code{xm} in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform to the within-patient slopes for the \code{~ 1 + x} model in the two data sets with the smallest between-patient variation. For data with a small between-patient variation, fixed-effects predictions for the \code{~ 1 + x} model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variation.  With data whose between-patient variation is relatively large, predictions based on the model with a common intercept and slope for all clusters are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

We therefore anticipate (and show later in this section) that case-based cross-validation may prefer the intercept-only model, \code{~ 1}  to the larger \code{~ 1 + x} model when the between-cluster variation is relatively small, but that cluster-based cross-validation will prefer the latter to the former. We will discover that case-based cross-validation prefers the \code{~ 1 + x} model to the \code{~ 1} model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' data set.

The third model, \code{~ 1 + x + xm}, includes a contextual effect of \code{x}---that is, the cluster mean \code{xm}---along with \code{x}, the intercept in the fixed-effect part of the model, and a random intercept. This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | patient)}, which is the model that generated the data.  The fit of the mixed model \code{~ 1 + x + xm} is consequently similar to that of a fixed-effects-only model with \code{x} and a categorical predictor for individual patients (i.e., \code{y ~ factor(patient) + x}, treating patients as a factor, and not shown here).

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only.[^cv-list] The results for all data sets, using both cluster-based and case-based cross-validation, are assembled in Figure \ref{fig:cross-validation-data-plot}.

@@

```{r cross-validation-data,include=FALSE,cache=TRUE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "75%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variation is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models (in Figure \ref{fig:plot-fits-combined} on page \pageref{fig:plot-fits-combined}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

As @HastieTibshiraniFriedman:2009 [Sec. 7.10.2: "The Wrong and Right Way to Do Cross-validation"] explain, if the whole data are used to specify or fine-tune a statistical model, then subsequent cross-validation of the model is intrinsically misleading, because the model is selected to fit the whole data, including the part of the data that remains when each fold is removed.  Statistical modeling is partly a craft, and one could imagine applying that craft independently to successive partial data sets, each with a fold removed. The resulting procedure would be tedious, though possibly worth the effort, but it would also be difficult to realize in practice: After all, we can hardly erase our memory of statistical modeling choices between analyzing partial data sets. Alternatively, if we're able to automate the process of model specification, then we can more realistically apply CV mechanically.

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore-scatterplot-matrix}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package); in each panel, the solid line shows the linear least-squares fit and the broken line is for a nonparametric regression:

```{r Auto-explore-scatterplot-matrix}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data."
Auto$cylinders <- factor(Auto$cylinders,
                         labels = c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels = c("AMER", "EUR", "JAP"))
rownames(Auto) <- make.names(Auto$name, unique = TRUE)
Auto$name <- NULL

@@

A comment before we proceed: \code{origin} is clearly categorical and so converting it to a factor is natural, but we could imagine treating \code{cylinders} and \code{year} as numeric predictors. There are, however, only 5 distinct values of \code{cylinders}, ranging from 3 to 8, but cars with 3 or 5 cylinders are rare, and none of the cars has 7 cylinders. There are similarly only 13 distinct years between 1970 and 1982 in the data, and the relationship between \code{mpg} and \code{year} is difficult to characterize.[^year] It's apparent that most of the numeric variables are positively skewed and that many of the pairwise relationships among them are nonlinear.

@@

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:

```{r Auto-explore-cr-plots}
#| out.width = "95%",
#| fig.height = 5.5,
#| fig.align = 'center',
#| fig.cap = "Component + residual plots for the working model fit to the \\code{Auto} data."
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch = ".", ylab = "C+R", las = 2)
```

The component + residual ("C+R") plots in Figure \ref{fig:Auto-explore-cr-plots}, produced by the \code{crPlots()} function in the \pkg{car} package, clearly reveal the inadequacy of the model. The broken blue lines in the C+R plots show linear least-squares fits, while the solid magenta lines are nonparametric-regression smooths.

@@

The \code{powerTransform()} function in the \pkg{car} package transforms variables towards multivariate normality by a generalization of Box and Cox's maximum-likelihood-like approach [@BoxCox:1964]. Several "families" of power transformations can be used, including the original Box-Cox family (which is the default), simple powers (and roots), and two adaptations of the Box-Cox family to data that may include negative values and zeros: the Box-Cox-with-negatives family and the Yeo-Johnson family; see @Weisberg:2014 [Chap. 8] and @FoxWeisberg:2019 [Chap. 3] for details. We proceed to transform the numeric predictors in the \code{Auto} regression towards multivariate normality: 
```{r Auto-transform}
num.predictors <- c("displacement", "horsepower", "weight", 
                    "acceleration")
tr.x <- powerTransform(Auto[, num.predictors])
summary(tr.x)
```
We then apply the (rounded) transformations---all, as it turns out, logs (i.e., "zeroth" powers)---to the data and re-estimate the model:
```{r Auto-with-transformed-predictors}
A <- Auto
powers <- tr.x$roundlam
for (pred in num.predictors){
  A[, pred] <- bcPower(A[, pred], lambda = powers[pred])
}
m <- update(m.auto, data = A)
```

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore-scatterplot-matrix}, on page \pageref{fig:Auto-explore-scatterplot-matrix}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "95%",
#| fig.align = 'center',
#| fig.height = 5.5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data."
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread = FALSE), data = A, pch = ".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:

```{r Auto-transformed-cr-plots}
#| out.width = "95%",
#| fig.align = 'center',
#| fig.height = 5.5,
#| fig.cap = "Component + residual plots for the regression model fit to the transformed \\code{Auto} data."
crPlots(m, pch = ".", ylab = "C+R", las = 2)
```

@@

* \code{data}, the data set to which the model is fit.
* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
* Arguments required by the model-selection function: the starting \code{working.model} (here, \code{m.auto}) for transformation and predictor selection; the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation; and \code{AIC = FALSE}, which specifies use of the BIC for model selection.
